Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
finished editing 3.2

The atomic specification for \texttt{insert} method is
\begin{mathpar}
	{\color{blue}
		m.\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
\wm{Consider skipping this paragraph if we need space.} Similar to the \texttt{traverse} specification, the \texttt{insert} function's specification is composed of two components: local and public pre- and postconditions. The local pre- and postconditions, containing $\nodeboxrep \ \texttt{r}$, describe the binary search tree's concrete state with the root node $\texttt{r}$, which points to the node's lock and lock invariant assertion. On the other hand, the public pre- and postconditions are represented by the predicate $\treerep \ m$, illustrating the binary search tree's abstract state as a collection of ghost states, with each state providing information about an individual node in the tree. Here, $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively.

Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template, annotated using separation logic.
The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{r}$. We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{proof_lock_insert}). After the lock for the root node is acquired (line 5 in Figure \ref{proof_lock_insert}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$. We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
		\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
		\right\rangle
	}
	\\ 
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
		\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
			\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
		\end{array}
		\ \Bigg| \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
In the $\mathsf{then}$ branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree. In this scenario, we update the node's value while maintaining the rest of the tree structure (line 9 in Figure \ref{proof_lock_insert}). Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 14-15 in Figure \ref{proof_lock_insert}). We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree. 
We can then release the lock (line 21 in Figure \ref{proof_lock_insert}) return the node's contents to the shared state. %Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
%\wm{Do we ever define $\treerep$?}

>>>>

The insert function uses this specification of \lstinline{traverse} to update the state of the data structure. The atomic specification for \texttt{insert} is
\begin{mathpar}
	{\color{blue}
		m.\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We can then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree.
We can then release the lock held by \lstinline{traverse} and deallocate the \lstinline{pn} structure.
